Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2020-02-15 article_springer/article_springer.Rmd
Peter's part of addressing referee comments

- name: "Peter Straka"
  affiliation: c
  email: "p.straka@unsw.edu.au"

@@

```{r thresholdedBursty, message=FALSE, fig.height=6, fig.width=7,out.width='70%',fig.align='center',fig.cap="\\label{fig:thresholdedBursty} Exceedances (red) and times until Exceedance (durations between blue crosses) for a given threshold $\\ell$ (dashed line). Upper picture: Simulated data with stable distributed waiting times. Lower picture: Solar flares during 1982."}
tail <- params$tail
n <- params$n 
sigma <- (cos(pi*tail/2))^(1/tail)
times <- cumsum(stabledist::rstable(n, tail, 1, sigma, pm=1))
magnitudes <- extRemes::revd(n, scale = 1, shape = 0)
sim_ctre <- ctre(data.frame(times, magnitudes)) 
p1<-plot(sim_ctre, p = 0.01, main = "Simulated MRP") 
p2<-flares %>% ctre() %>% plot(p = 0.02, log = 'y', main = "HXRBS data") 
grid.arrange(p1,p2)

In this article, we restrict ourselves to the uncoupled case, where $W$ and $J$ are independent. Then the two sequences $X(\ell, n)_{n \in \mathbb N}$ and $T(\ell, n)_{n \in \mathbb N}$ are independent as well. To see why, note that $X(\ell)$ is, in distribution, simply equal to $J | J > \ell$, independent of any waiting time $W_k$.  
Figure \ref{fig:thresholdedBursty} shows a simulated dataset in the top panel, where $W$ has a stable distribution with tail parameter $\beta =$ `r params$tail` (and skewness $1$ and location $0$), and where $J$ is from a standard Gumbel distribution. In the bottom panel, we plot a time series of solar flare intensities derived from a NASA dataset [@HXRBS] which we will later examine more closely (see Section 7). Clearly, the simulated data exhibit long intervals 
_without any_ events, whereas in the real-world dataset events appear continuously. The threshold exceedances, however, appear to have visually similar statistical behaviour in both models. Observations below a threshold are commonly discarded in Extreme Value Theory (POT approach); likewise, the CTRE model interprets these observations as noise and discards them.

@@

In this section we state and prove the key theorem, see below. For an accessible introduction to regular variation and stable limit theorems, we recommend the book by @MeerschaertSikorskii.

@@

: Let the waiting times $J_k$ be in the domain of attraction of a positively skewed sum-stable law with stability parameter $0 < \beta < 1$; more precisely,
\begin{align} \label{eq:stability}
\frac{W_1 + \ldots + W_n}{b(n)} \overset{d}{\longrightarrow} D, 
\quad n \to \infty
\end{align}
for a function $b(n)$ which is regularly varying at $\infty$ with parameter $1/\beta$, and where $\mathbf E[\exp(-sD)] = \exp(-s^\beta)$. Write $p := \mathbf P(J > \ell)$. Then the weak convergence
$$
\frac{T(\ell)} {b(1/p)} \to W_\beta \quad \text{ as } \quad \ell \uparrow x_R
$$
holds, where the Mittag-Leffler random variable $W_\beta$ is defined on 
the positive real numbers via 
$$
\mathbf E[\exp(-sW_\beta)] = \frac{1}{1+s^\beta}.
$$

*Proof of Theorem:* We interpret the threshold crossing time $T(\ell)$ as the hitting time of the 
underlying CTRM (Continuous Time Random Maxima) or "max-renewal process", 
and then utilize a result by @MeerschaertStoev08. 
The running maximum process is defined as 
$$
M(c) := J_1 \vee \ldots \vee J_{\lfloor c \rfloor},
$$
and since we assume that the $J_k$ have a continuous distribution, there exist 
norming functions $a(c)$ and $d(c)$ such that 
$$
\mathbf P\left[ \frac{M(c) - d(c)}{a(c)} \le \ell^* \right] 
\longrightarrow F(\ell^*), \quad t \to \infty
$$
where $F$ is a generalized extreme value distribution, and $\ell^*$ is any 
value from the support of $F$. 
The CTRM process is then defined via 
$$
V(t) = M(N(t)), \quad t \ge 0
$$
where $N(t)$ is the renewal process associated with the waiting times 
$W_k$: 
$$
N(t) = \max\{n: W_1 + \ldots + W_n \le t\}.
$$
Now a key observation is that
$$
T(\ell) = \inf\{t: V(t) > \ell\}, 
$$
and that 
$$
T(\ell) > t \quad \text{ if and only if } \quad V(t) \le \ell.
$$
By [Theorem 3.1, @MeerschaertStoev08], we have the stochastic process 
convergence 
$$
\frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\stackrel{d}{\longrightarrow} Y(t), \quad t > 0.
$$
where $Y(t)$ is a time-changed ("subordinated") extremal process, and where 
$\tilde b(c)$ is a regularly varying norming function which is
_inverse_ to $b(c)$, in the sense that 
$b(\tilde b(c)) \sim c \sim \tilde b(b(c))$.

@@

For a scale parameter $\sigma > 0$, we write ${\rm ML}(\beta, \sigma)$ for the distribution of $\sigma W_\beta$. The Mittag-Leffler distribution with parameter $\beta \in (0,1]$ is 
a heavy-tailed positive distribution for $\beta < 1$, with infinite mean. 
However, as $\beta \uparrow 1$, ${\rm ML}(\beta, \sigma)$ converges 
weakly to the exponential distribution ${\rm Exp}(\sigma)$ with mean $\sigma$.
This means that although its moments are all infinite, the Mittag-Leffler
distribution may (if $\beta$ is close to 1) be indistinguishable from the 
exponential distribution, for the purposes of applied statistics. 
For a detailed reference on the Mittag-Leffler distribution, see e.g. 
@Haubold11, and for algorithms, see e.g. the R package `MittagLeffleR`
[@MittagLeffleR].

@@

@@

Due to the form of the density function of a Mittag-Leffler distribution there exists no closed form for the Maximum Likelihood estimator (MLE). In the R Package `MittagLeffleR` [@MittagLeffleR], Maximum Likelihood estimation is implemented via numerical optimization. The MLE slightly outperforms the log-moment estimator regarding bias and RMSE for big enough sample sizes, but is extremely computational intensive. Figure \ref{fig:MSE} shows that both estimators perform well, even for small sample sizes.

```{r MSE, echo=FALSE,warning=FALSE, fig.height=7,fig.width=8,fig.align="center", fig.cap="\\label{fig:MSE} MSE for the estimation of tail (left column) and scale (right column) parameters via log-moment estimator and mle of a Mittag-Leffler sample with varying sample size n=25, 100, 200, varying tails on the x-axis and fixed scale equal to one.", out.width = '90%'}
load(here::here("article_springer\\article_springer_files\\ML_estim_simu","tbl_ges.RData"))
tbl_ges %>% 
  filter(n == "25" || n == "100" || n=="200") %>%
  filter(true.scale == "1") %>% 
  ggplot(aes(true.tail, mse, colour = method, group=method)) + 
  geom_line() + geom_point() + 
  facet_wrap(c("n","param"), scales = "free", ncol = 2) + 
  scale_color_manual(values=c("red", "blue"))
```

Since the Mittag-Leffler distribution is heavy-tailed, many researchers would intuitively give the highest importance to the tail behaviour of the distribution. Of course, one can also use established tail exponent estimators for the estimation of the parameter $\beta$ such as the Hill estimator [@hill1975simple] based on the $r+1$ upper order statistics
\begin{align}\label{eq:hill}
H_{r,n}=\left[ \frac{1}{r} \sum_{i=0}^{r} \log \frac{X_{(i)}}{X_{(r+1)}}\right]^{-1}, 
\end{align}
where $X_{(1)} \geq X_{(2)} \geq \ldots \geq X_{(n)}$ denote the order statistics in decreasing order of a sequence $X_1,\ldots,X_n$. However, these methods are statistically less efficient since they only use a portion of the information contained in the data, as mentioned by @kozubowski2001. Moreover, the hill estimator requires a tuning parameter, denoted with $r$ in \eqref{eq:hill}. Additionaly, the estimator only performs reliably well for distributions close to Pareto [see @Resnick97]. Figure \ref{fig:Hillplots} shows Hill plots for Mittag-Leffler simulated data, with varying sample sizes and tails. To deduce the correct tail parameter estimates from these plots is virtually impossible. To estimate the tail parameter of a Mittag-Leffler distribution with the hill estimator, the sample size has to be much larger and one has to choose a large tuning parameter $r$. In case that only events above a high threshold are recorded, even $n=1000$ events are unrealistic. Furthermore the Hill estimator of course completely fails if the inter-arrival times are exponential distributed and not heavy-tailed. We will comeback to the problems with the hill estimtator in Section \ref{Simulationstudy}.

```{r Hillplots, echo=FALSE,warning=FALSE, fig.align="center", fig.cap="\\label{fig:Hillplots} Hillplots for simulated Mittag-Leffler data with true tail 0.8 and sample size 200 (left) and 1000 (right), with number of upper order statics r on which the hill estimator is based on the x-axis. ", out.width = '90%'}
hill_estim<-function(k,data){
  n<-length(data)
  data<-sort(data)
  data_thinned<-data[(n-k):n]
  r<-1/k * sum(log(data_thinned/data_thinned[1]))
  return(1/r)
}

@@

data_ml_1<-rml(n=200,tail=0.8,scale=1)
p1<-hill_estim_plot(k_min=30,data=data_ml_1,true_tail=0.8)

data_ml_2<-rml(n=1000,tail=0.8,scale=1)
p2<-hill_estim_plot(k_min=30,data=data_ml_2,true_tail=0.8)
grid.arrange(p1,p2,ncol=2)
```

@@

```{r LRT_power, echo=FALSE,warning=FALSE, fig.align="center", fig.cap="\\label{Fig:LRT} Power for bootstrapped LRT for different sample sizes , varying tails and scale parameter equal to 1.",out.width='90%'}
load(here::here("article_springer","article_springer_files","ML_LRT_simu","estimates_df_LRT_boot1_01.RData"))
df_LRT_boot_01 %>% filter(n=="20" || n=="50" || n=="100" || n=="200") %>% ggplot(aes(x=true.tail,y=power)) +  geom_line(aes(y=power,group=factor(n),color=factor(n),linetype=factor(n)),size=0.5)+
  scale_linetype_manual(values = c("solid","dashed","twodash","dotted"),name  ="Sample size")+
   labs(y="Power",x= "True tail parameter")+
  scale_color_manual(values=c("red", "blue","darkgreen","black"), name  ="Sample size")
```

@@

1. For a range of thresholds $\ell$ near the largest order statistics, extract datasets of exceedance times $\{T(\ell, i)\}_i$.

2. For each choice of threshold $\ell$, fit a Mittag-Leffler distribution to the resulting dataset
  $\{T(\ell, i)\}_i$.  This results in the estimates $\{\hat\beta(\ell)\}_\ell$ and $\{\hat \sigma(\ell)\}_\ell$.

3. Plot $\ell$ vs.\ $\hat \beta(\ell)$. As $\ell$ increases towards $x_R$, 
  $\hat \beta(\ell)$ *stabilizes* around a constant $\hat \beta$. Use $\hat \beta$ as an estimate for the tail      parameter $\beta$ of the Mittag-Leffler distribution of exceedance times.

4. Approximate $p \approx |\{k: J_k > \ell\}| / n$. 
  Recall that $b(c)$ is regularly varying with parameter $1/\beta$, and 
  hence has the representation $b(c) = L(c) c^{1/\beta}$ for some 
  slowly varying function $L(c)$. 
  Assuming that the variation of $L(c)$ is negligible, we hence 
  plot $\ell$ vs.\ $p^{1/\hat \beta} \hat \sigma(\ell)$. 
  Again as $\ell$ increases towards $x_R$, 
  $p^{1/\hat \beta} \hat \sigma(\ell)$ is expected to stabilize around a constant 
  $\hat \sigma_0$. 
  We then use $p^{-1/\hat \beta} \hat \sigma_0$ as an estimate of the scale 
  parameter of the Mittag-Leffler distribution of exceedance times for
  the level $\ell$.

The above approach benefits from the following
practical adjustments (compare with Figure \ref{fig:flares}):

* We choose $\ell$ from the order statistics, i.e. $\ell$ is the $k$-th
  largest of the observations $X_j$, where $k$ runs from 
  $k_\text{min}, k_\text{min} + 1, \ldots, k_\text{max}$. 
  The datasets are then of length $k-1$.

* We use $k$ rather than $\ell$ for the horizontal axis of our plots.

* In Step 4, rather than plotting $p^{1/\hat \beta} \hat \sigma(\ell)$
  we plot $k^{1/\hat \beta} \hat \sigma(\ell)$. This changes 
  $\hat \sigma_0$ by the multiplicative constant $n^{1/\hat \beta}$, 
  but has the advantage that
  $\hat \sigma_0$ does not change if one pre-processes the data by 
  removing all observations below a certain threshold.

The estimates $\hat \beta$ and $\hat \sigma_0$ give an estimate of the 
distribution of exceedance times, dependent on the threshold $\ell$:
\begin{align*}
T(\ell) \sim {\rm ML}(\hat \beta, k^{-1/\hat \beta} \hat \sigma_0),
\end{align*}
where $\ell$ equals the $k$-th order statistic.
For quick estimates of the Mittag-Leffler parameters we have used 
the method of log-transformed moments [@Cahoy2013].
We have verified the validity of our estimation algorithm  via simulations, 
see Section \ref{Simulationstudy}.

# Simulation study{#Simulationstudy}

@@

In order to have exact analytical values available for $\beta$ and $\sigma_0$, a distribution for $W_k$ needs to be chosen for which $b(n)$ from \eqref{eq:stability} is known. For (i) we choose $W_k \stackrel{d}{=} D$, where $D$ is as in \eqref{eq:stability}, then due to the stability property we have the *equality* of distribution
$W_1 + \ldots + W_n \stackrel{d}{=} b(n) D$, 
for $b(n) = n^{1/\beta}$. 
Using the parametrisation of @SamorodnitskyTaqqu, a few lines of 
calculation [see e.g. the vignette on parametrisation in @MittagLeffleR]
show that $D$ must have the stable distribution 
$S_\beta(\cos(\pi \beta/2)^{1/\beta}, +1, 0)$, which is 
implemented in the R package `stabledist` by @stabledist. By the Theorem, the distribution of $T(\ell)$ is approximately 
$$
{\rm ML}(\beta, p^{-1/\beta}) 
= {\rm ML}(\beta, k^{-1/\beta} n^{1/\beta}),
$$
which means 
$\sigma_0 = n^{1/\beta}$. In the Pareto example we choose $P(W>t)=Ct^{-\beta}$ with $C=(1/\Gamma(1-\beta))^{1/\beta}$. We have chosen $\beta=0.8$ in the stable as well as in the Pareto case. As third example we choose exponential distributed waiting times with a rate of $1$. Figure \ref{Fig:TailSimu} shows the stability plots for the estimation of the tail parameter. Therefore, $\hat \beta(k)$ vs. $k$ is plotted for each of the $m=100$ simulation runs (grey thin lines). In the left column one can see the estimates based on the log-moment estimator, the stability plots in the middle are based on the maximum-likelihood estimator and on the right hand on the hill estimator. Recall that $k$ is the index of the order statistics of $J_k$ at which the threshold $\ell$ is placed. Note, that the hill estimator is here based on the waiting times, not on the inter-exceedance times, hence the estimation is based on much more data. If one would base the hill-estimation only on the inter-exceedance times, the sample size would be too small and it would be impossible to deduce an estimate for the tail parameter from the plots (compare Figure \ref{fig:Hillplots}). The $k$ on the $x-$axis is in the plots for the hill estimator the $r$ which defines on which upper orders stats the hill estimator is based on (compare \eqref{eq:hill}). The bigger sample size results in the middle in a lower bias. But, one can also see, that there is much more variance between the stability plots for the different simulation runs in case of the hill estimator compared to the other two estimators. The black line is the mean value of the point estimators for each $k$ resp. $r$. However, if one wants to get a single estimate via such a stability plot, it would be much more difficult to deduce it from a hill plot compared to the other two estimators. In case that one has only thresholded data and hence a much smaller simple size, the hill estimator gets useless (see Section \ref{sec:ML}). Another advantage of the other estimators compared to the hill estimator is, that in the case one has exponential waiting times, the log-moment as well as the mle can detect this by getting an estimate close to $1$. The hill estimators is just a tail parameter estimator for Pareto like tails and hence of course fails in case of exponential distributed inter-exceedance times. Figure \ref{Fig:ScaleSimu} shows the stability plots for the different estimators for the scale parameter. Here of course only for the log-moment (on the right hand)  and the mle (on the left hand), since the hill estimator is just a tail parameter estimate. Here again both estimators show good performance. The stability plots in the mle case are a little bit closer to the true value, except in the exponetial case. Therefore, the mle is much more computational intensive.

```{r TailSimuplots, message=FALSE,echo=FALSE,fig.height=7,fig.align="center",fig.cap="\\label{Fig:TailSimu} Stability Plots for m=100 simulation runs for stable distributed waiting times with a tail parameter of 0.8 (top row), Pareto distributed waiting times with tail parameter 0.8 (middle row) and exponential distributed waiting times (lower row). Left column: log-moment estimator, middle column: MLE, right column: Hill estimator. The grey thin lines are the stability plots for the different simulation runs and the dark lines are their means. The red dotted line shows the true tail parameter. ", fig.show='hold', out.width = '100%'}

path<-here::here("article_springer/article_springer_files/Simulation_results")
files<-list.files(path)
file_dirs<-paste(path,"/",files,sep="")
invisible(sapply(file_dirs,FUN=load,envir=.GlobalEnv,verbose=FALSE))

@@

Before we apply the POT approach described in Section 5 to the solar flare data, we first have to check if all model assumptions are full-filled. The CTRE model is based on three main assumptions, which are repeated below. For each assumption, we suggest one means of checking if it holds:

>>>>

- name: "Peter Straka"
  affiliation: c
  email: "straka.ps@gmail.com"

@@

```{r thresholdedBursty, message=FALSE, fig.height=6, fig.width=7,out.width='70%',fig.align='center',fig.cap="\\label{fig:thresholdedBursty} Exceedances (red) and times until Exceedance (durations between blue crosses) for a given threshold $\\ell$ (dashed line). Upper picture: Simulated data with stable distributed waiting times. Lower picture: Solar flares during 1982."}
tail <- params$tail
n <- params$n 
sigma <- (cos(pi*tail/2))^(1/tail)
times <- cumsum(stabledist::rstable(n, tail, 1, sigma, pm=1))
magnitudes <- extRemes::revd(n, scale = 1, shape = 0)
sim_ctre <- ctre(data.frame(times, magnitudes)) 
p1<-plot(sim_ctre, p = 0.01, main = "Simulated MRP") + 
  ggtitle("Simulated MRP")
p2<-flares %>% ctre() %>% plot(p = 0.02, log = 'y', main = "HXRBS data") +
  ggtitle("HXRBS data")
grid.arrange(p1,p2)
```

In this article, we restrict ourselves to the uncoupled case, where $W$ and $J$ are independent. Then the two sequences $X(\ell, n)_{n \in \mathbb N}$ and $T(\ell, n)_{n \in \mathbb N}$ are independent as well. To see why, note that $X(\ell)$ is, in distribution, simply equal to $J - \ell | J > \ell$, independent of any waiting time $W_k$.  
Figure \ref{fig:thresholdedBursty} shows a simulated dataset in the top panel, where $W$ has a stable distribution with tail parameter $\beta =$ `r params$tail` (and skewness $1$ and location $0$), and where $J$ is from a standard Gumbel distribution. In the bottom panel, we plot a time series of solar flare intensities derived from a NASA dataset [@HXRBS] which we will later examine more closely (see Section 7). Clearly, the simulated data exhibit long intervals 
_without any_ events, whereas in the real-world dataset events appear continuously. The threshold exceedances, however, appear to have visually similar statistical behaviour in both models. Observations below a threshold are commonly discarded in Extreme Value Theory (POT approach); likewise, the CTRE model interprets these observations as noise and discards them.

@@

In this section we state and prove the key theorem. For an accessible introduction to regular variation and stable limit theorems, we recommend the book by @MeerschaertSikorskii.

@@

: Let the waiting times $W_k$ be in the domain of attraction of a positively skewed sum-stable law with stability parameter $0 < \beta < 1$; more precisely,
\begin{align} \label{eq:stability}
\frac{W_1 + \ldots + W_n}{b(n)} \overset{d}{\longrightarrow} D, 
\quad n \to \infty
\end{align}
for a function $b(n)$ which is regularly varying at $\infty$ with parameter $1/\beta$, and where $\mathbf E[\exp(-sD)] = \exp(-s^\beta)$. Write $p := \mathbf P(J > \ell)$. Then the weak convergence
$$
\frac{T(\ell)} {b(1/p)} \to W_\beta \quad \text{ as } \quad \ell \uparrow x_R
$$
holds, where the Mittag-Leffler random variable $W_\beta$ is defined on 
the positive real numbers via 
$$
\mathbf E[\exp(-sW_\beta)] = \frac{1}{1+s^\beta}.
$$

*Proof of Theorem:* We interpret the threshold crossing time $T(\ell)$ as the hitting time of the 
underlying CTRM (Continuous Time Random Maxima) or "max-renewal process", 
and then utilize a result by @MeerschaertStoev08. 
The running maximum process is defined as 
$$
M(c) := J_1 \vee \ldots \vee J_{\lfloor c \rfloor},
$$
and since we assume that the $J_k$ have a continuous distribution, there exist 
norming functions $a(c)$ and $d(c)$ such that 
$$
\mathbf P\left[ \frac{M(c) - d(c)}{a(c)} \le \ell^* \right] 
\longrightarrow F(\ell^*), \quad t \to \infty
$$
where $F$ is a generalized extreme value distribution, and $\ell^*$ is any 
value from the support of $F$. 
The CTRM process is then defined via 
$$
V(t) = M(N(t)), \quad t \ge 0
$$
where $N(t)$ is the renewal process associated with the waiting times 
$W_k$: 
$$
N(t) = \max\{n: W_1 + \ldots + W_n \le t\}.
$$
Now a key observation is that
$$
T(\ell) = \inf\{t: V(t) > \ell\}, 
$$
and that 
$$
T(\ell) > t \quad \text{ if and only if } \quad V(t) \le \ell.
$$
By [Theorem 3.1, @MeerschaertStoev08], we have the stochastic process 
convergence 
$$
\frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\stackrel{d}{\longrightarrow} Y(t), \quad t > 0.
$$
where $Y(t)$ is a time-changed ("subordinated") extremal process, and where 
$\tilde b(c)$ is a regularly varying norming function which is
_inverse_ to $b(c)$, in the sense that 
$b(\tilde b(c)) \sim c \sim \tilde b(b(c))$ [@BinghamGoldieTeugels].

@@

For a scale parameter $\sigma > 0$, we write ${\rm ML}(\beta, \sigma)$ for the distribution of $\sigma W_\beta$. The Mittag-Leffler distribution with parameter $\beta \in (0,1]$ is 
a heavy-tailed positive distribution for $\beta < 1$, with infinite mean. 
However, as $\beta \uparrow 1$, ${\rm ML}(\beta, \sigma)$ converges 
weakly to the exponential distribution ${\rm Exp}(\sigma)$ with mean $\sigma$.
This means that although its moments are all infinite, the Mittag-Leffler
distribution may (if $\beta$ is close to 1) be indistinguishable from the 
exponential distribution, for the purposes of applied statistics.

We caution the reader that, somewhat confusingly, there is another distribution
called the "light-tailed" Mittag-Leffler distribution.  This is in fact the 
limiting distribution of the renewal process $N(t)$ above (see @limitCTRW). 
For a detailed reference on the Mittag-Leffler distribution, see e.g. 
@Haubold11, and for algorithms, see e.g. the R package `MittagLeffleR`
[@MittagLeffleR].

@@

: When $0 < \beta < 1$, the renewal process $N(t)$ is _not stationary_, 
and hence the results by @Hsing88 on the exceedances of stationary sequences
do not apply.

@@

Due to the form of the density function of a Mittag-Leffler distribution there exists no closed form for the Maximum Likelihood estimator (MLE). In the R Package `MittagLeffleR` [@MittagLeffleR], Maximum Likelihood estimation is implemented via numerical optimization. The MLE slightly outperforms the log-moment estimator regarding bias and RMSE for big enough sample sizes, but is extremely computationally intensive. Figure \ref{fig:MSE} shows that both estimators perform well, even for small sample sizes.

```{r MSE, echo=FALSE,warning=FALSE, fig.height=7,fig.width=8,fig.align="center", fig.cap="\\label{fig:MSE} MSE for the estimation of tail (left column) and scale (right column) parameters via log-moment estimator and MLE of a Mittag-Leffler sample with varying sample size n=25, 100, 200, varying tails on the x-axis and fixed scale equal to one.", out.width = '90%'}
load(here::here("article_springer", "article_springer_files", "ML_estim_simu","tbl_ges.RData"))
tbl_ges %>% 
  filter(n == "25" || n == "100" || n=="200") %>%
  filter(true.scale == "1") %>% 
  ggplot(aes(true.tail, mse, colour = method, group=method)) + 
  geom_line() + geom_point() + 
  facet_wrap(c("n","param"), scales = "free", ncol = 2) + 
  scale_color_manual(values=c("red", "blue"))
```

Since the Mittag-Leffler distribution is heavy-tailed, many researchers would intuitively give the highest importance to the tail behaviour of the distribution. Of course, one can also use established tail exponent estimators for the estimation of the parameter $\beta$ such as the Hill estimator [@hill1975simple] based on the $r+1$ upper order statistics
\begin{align}\label{eq:Hill}
H_{r,n}=\left[ \frac{1}{r} \sum_{i=0}^{r} \log \frac{X_{(i)}}{X_{(r+1)}}\right]^{-1}, 
\end{align}
where $X_{(1)} \geq X_{(2)} \geq \ldots \geq X_{(n)}$ denote the order statistics in decreasing order of a sequence $X_1,\ldots,X_n$. However, these methods are statistically less efficient since they only use a portion of the information contained in the data, as mentioned by @kozubowski2001. Moreover, the Hill estimator requires a tuning parameter, denoted with $r$ in \eqref{eq:Hill}. Additionaly, the estimator only performs reliably well for distributions close to Pareto [see @Resnick97]. Figure \ref{fig:Hillplots} shows Hill plots for Mittag-Leffler simulated data, with varying sample sizes and tails. To deduce the correct tail parameter estimates from these plots is virtually impossible. To estimate the tail parameter of a Mittag-Leffler distribution with the Hill estimator, the sample size has to be much larger and one has to choose a large tuning parameter $r$. In case that only events above a high threshold are recorded, even $n=1000$ events are unrealistic. Furthermore the Hill estimator of course completely fails if the inter-arrival times are exponentially distributed and not heavy-tailed. We will come back to the problems with the Hill estimtator in Section \ref{Simulationstudy}.

```{r Hillplots, echo=FALSE,warning=FALSE, fig.align="center", fig.cap="\\label{fig:Hillplots} Hillplots for simulated Mittag-Leffler data with true tail 0.8 and sample size 200 (left) and 1000 (right), with number of upper order statics r on which the Hill estimator is based on the x-axis. ", out.width = '90%'}
hill_estim<-function(k,data){
  n<-length(data)
  data<-sort(data)
  data_thinned<-data[(n-k):n]
  r<-1/k * sum(log(data_thinned/data_thinned[1]))
  return(1/r)
}

@@

data_ml_1<-rml(n=200,tail=0.8,scale=1)
p1<-hill_estim_plot(k_min=30,data=data_ml_1,true_tail=0.8) +
  ggtitle("Hill plot, Mittag-Leffler data, n=200")

data_ml_2<-rml(n=1000,tail=0.8,scale=1)
p2<-hill_estim_plot(k_min=30,data=data_ml_2,true_tail=0.8) +
  ggtitle("Hill plot, Mittag-Leffler data, n=1000")

grid.arrange(p1,p2,ncol=2)
```

@@

```{r LRT_power, echo=FALSE,warning=FALSE, fig.align="center", fig.cap="\\label{Fig:LRT} Power for bootstrapped LRT for different sample sizes, varying tails and scale parameter equal to 1.", out.width='90%'}
load(here::here("article_springer","article_springer_files","ML_LRT_simu","estimates_df_LRT_boot1_01.RData"))
df_LRT_boot_01 %>% filter(n=="20" || n=="50" || n=="100" || n=="200") %>% ggplot(aes(x=true.tail,y=power)) +  geom_line(aes(y=power,group=factor(n),color=factor(n),linetype=factor(n)),size=0.5)+
  scale_linetype_manual(values = c("solid","dashed","twodash","dotted"),name  ="Sample size")+
   labs(y="Power",x= "True tail parameter")+
  scale_color_manual(values=c("red", "blue","darkgreen","black"), name  ="Sample size") +
  ggtitle("Power of Likelihood-Ratio Test")
```

@@

1. Extract the $K$ largest order statistics 
(i.e. the $K$ largest values, where e.g. $K = 300$) 
$X_{(1)}, \ldots, X_{(K)}$ together with their timestamps $T_{(1)}, \ldots, T_{(K)}$.

2. Choose e.g. $K_0 = 5$, and for each $k$ ranging from $K_0 +1$ to $K + 1$: 
    a) extract the set $\mathcal T_k$ of exceedance times of the threshold $X_{(k)}$
    b) fit a Mittag-Leffler distribution to $\mathcal T_k$, resulting in the parameter
    estimates $\hat\beta_k$ and $\hat \sigma_k$.

3. Plot $k$ vs.\ $\hat \beta_k$. As $k \downarrow K_0$,
  $\hat \beta_k$ *stabilizes* around a constant $\hat \beta$.

4. Plot $k$ vs.\ $k^{1/\hat \beta} \hat \sigma_k$. 
  Where the plot stabilizes as $k \downarrow K_0$, define the value 
  $\hat \sigma_0$.

The inferred values $\hat \beta$ and $\hat \sigma_0$ can then be interpreted as follows: 
Setting the threshold at $X_{(k)}$, the threshold exceedance times follow a 
Mittag-Leffler distribution with shape parameter $\hat \beta$ and scale parameter
$\hat \sigma_0 k^{-1/\beta}$.

To clarify Step 4: Recall that by the theorem, $\sigma_k / b(1/p)$ is expected to stabilize 
around a constant as $k \downarrow K_0$. Since $b$ is regularly varying with parameter
$1/\beta$, we have $b(1/p) = p^{-1/\beta} / L(1/p)$ for some slowly varying function $L$. 
Approximating $p \approx \hat p := k / n$, we have 
$$\text{const} \approx \sigma_k / b(1/p) = \sigma_k \times p^{1/\beta} L(1/p) \approx \sigma_k \times k^{1/\beta} n^{-1/\beta} L(n/k)$$
Assuming that the variation of $L(n/k)$ is minor, we can hence see that 
$\sigma_k k^{1/\beta}$ stabilizes.

For computationally efficient estimates of the Mittag-Leffler parameters we have used 
the method of log-transformed moments. This estimation method provides 
point estimates as well as confidence intervals based on sampling variance [@Cahoy2013], 
and has been implemented in the R software package
`MittagLeffleR` [@MittagLeffleR]. The stability plots for $\hat \beta$ and $\hat \sigma_0$
can be furnished with these confidence intervals, see e.g.\ Figure \ref{fig:flares}, 
to produce (non-simultaneous) confidence bands. 
These stability plots were produced with the R package `CTRE` [@CTRE]. 
We have verified the validity of our estimation algorithm  via simulations, 
see Section \ref{Simulationstudy}.

# Simulation Study {#Simulationstudy}

@@

In order to have exact analytical values available for $\beta$ and $\sigma_0$, a distribution for $W_k$ needs to be chosen for which $b(n)$ from \eqref{eq:stability} is known.

**Case (i):** For (i) we choose $W_k \stackrel{d}{=} D$, where $D$ is as in \eqref{eq:stability}, then due to the stability property we have the *equality* of distribution
$W_1 + \ldots + W_n \stackrel{d}{=} b(n) D$, 
for $b(n) = n^{1/\beta}$. 
Using the parametrisation of @SamorodnitskyTaqqu, a few lines of 
calculation [see e.g. the vignette on parametrisation in @MittagLeffleR]
show that $D$ must have the stable distribution 
$S_\beta(\cos(\pi \beta/2)^{1/\beta}, +1, 0)$, which is 
implemented in the R package `stabledist` by @stabledist. By the Theorem, the distribution of $T(\ell)$ is approximately 
$$
{\rm ML}(\beta, p^{-1/\beta}) 
= {\rm ML}(\beta, k^{-1/\beta} n^{1/\beta}),
$$
which means 
$\sigma_0 = n^{1/\beta}$.

**Case (ii):** In the Pareto example we choose $P(W>t)=Ct^{-\beta}$ with $C=(1/\Gamma(1-\beta))^{1/\beta}$. We have chosen $\beta=0.8$ in both cases (i) and (ii).

**Case (iii):** We choose exponentially distributed waiting times with a rate 
parameter of $1$.

Figure \ref{Fig:TailSimu} shows the graphical "stability plots" for the 
estimation of the tail parameter, where

+ rows correspond to cases (i), (ii) and (iii), and
+ columns correspond to estimators (log-moment estimator, maximum-likelihood, and Hill estimator).

We plot the tail parameter estimates $\hat \beta(k)$ against $k$ 
for each of the $m=100$ simulation runs. 
Thin grey lines represent individual simulation runs, 
and the thicker black line is their mean. 
Recall that $k$ is the index of the order statistics of $J_k$ at which the 
threshold $\ell$ is placed. 
Note that the Hill estimator is based on the waiting times, not on the 
inter-exceedance times, meaning that the estimation is based on more data. 
If one were to base the Hill-estimation only on the inter-exceedance times, 
the sample size would be too small and it would be impossible to deduce an 
estimate for the tail parameter from the plots (compare Figure \ref{fig:Hillplots}).

In the right-hand Hill-estimator column, the $k$ on the $x$-axis corresponds to
$r$ from \eqref{eq:Hill}.
The Hill estimator appears to be nicely unbiased on Pareto data,
but compared to the other two estimators, it has a higher variance, 
as can be seen from the wider spread of the simulated curves. 
This high variance means that that a single estimate is less reliable compared
to the other two estimators. 
This effect is amplified if one only has a thresholded, and thus smaller dataset
available, to the extent that the Hill estimates become severely flawed
(see Section \ref{sec:ML}).

Moreover, the bottom in \ref{Fig:TailSimu} shows clearly that the log-moment 
and maximum-likelihod estimators generalize to the exponential case ($\beta = 1$), 
whereas the Hill-estimator does not.

Finally, the performance of Log-Moment and Maximum-Likelihood estimators for the
scale parameter is shown in figure \ref{Fig:ScaleSimu}. 
Since there is no Hill-estimate of a scale parameter, there are only two columns. 
Both estimators show good performance, with a slight advantage for the 
Maximum-Likelihood estimator. This advantage is paid for by a higher computational 
cost.

```{r TailSimuplots, message=FALSE,echo=FALSE,fig.height=7,fig.align="center",fig.cap="\\label{Fig:TailSimu} Stability Plots for m=100 simulation runs for stable distributed waiting times with a tail parameter of 0.8 (top row), Pareto distributed waiting times with tail parameter 0.8 (middle row) and exponentially distributed waiting times (lower row). Left column: log-moment estimator, middle column: MLE, right column: Hill estimator. The grey thin lines are the stability plots for the different simulation runs and the dark lines are their means. The red dotted line shows the true tail parameter. ", fig.show='hold', out.width = '100%'}

path <-
  here::here("article_springer",
             "article_springer_files",
             "Simulation_results")
files <- list.files(path)
file_dirs <- paste(path, "/", files, sep = "")
invisible(sapply(
  file_dirs,
  FUN = load,
  envir = .GlobalEnv,
  verbose = FALSE
))

@@

Before we apply the POT approach described in Section 5 to the solar flare data, we first have to check if all model assumptions are fulfilled. The CTRE model is based on three main assumptions, which are repeated below. For each assumption, we suggest one means of checking if it holds:
